\documentclass[12pt]{article}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}

\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows.meta,arrows}

\newcommand{\tth}[1]{\footnote{\textcolor{blue}{#1}}}
\newcommand{\tbh}[1]{\textsc{\textbf{#1}}}

\title{\vspace{-2em}Technical Report: Exploring Automatic Model-Checking of the Ethereum specification\footnote{%
		The specifications, scripts, and experimental results can be accessed at: \url{https://github.com/freespek/ssf-mc}}}
\author{%
    Igor Konnov\thanks{This work was supported by Ethereum Foundation
    under grant FY24--1535.}\\ \footnotesize Independent Researcher%
    \and
    Jure Kukovec\footnotemark[2] \\ \footnotesize Independent Researcher%
    \and
    Thomas Pani\footnotemark[2] \\ \footnotesize Independent Researcher%
    \and
    Roberto Saltini\thanks{This work was supported by Ethereum Foundation
    under grant FY24--1536 and was partially done while at Consensys.} \\
    \footnotesize Independent Researcher%
    \and
    Thanh Hai Tran\footnotemark[3] \\ \footnotesize Independent Researcher%
}
\date{}

\input{preambule}

\begin{document}

\maketitle

\begin{abstract}%
We investigate automated model-checking of the Ethereum specification, focusing
on the \emph{Accountable Safety} property of the 3SF consensus protocol. We
select 3SF due to its relevance and the unique challenges it poses for formal
verification. Our primary tools are~\tlap{} for specification and the Apalache
model checker for verification.

Our formalization builds on the executable Python specification of 3SF\@. To
begin, we manually translate this specification into~\tlap{}, revealing
significant combinatorial complexity in the definition of Accountable Safety.
To address these challenges, we introduce several layers of manual abstraction:
(1) replacing recursion with folds, (2) substituting abstract graphs with
integers, and (3) decomposing chain configurations.
To cross-validate our results, we develop alternative encodings in SMT (CVC5)
and Alloy.

Despite the inherent complexity, our results demonstrate that exhaustive
verification of Accountable Safety is feasible for small instances --
supporting up to 7 checkpoints and 24 validator votes. Moreover, no violations
of Accountable Safety are observed, even in slightly larger configurations.
Beyond these findings, our study highlights the importance of manual
abstraction and domain expertise in enhancing model-checking efficiency and
showcases the flexibility of~\tlap{} for managing intricate specifications.%
\end{abstract}%
\newpage%

\setcounter{tocdepth}{2}  % hide anything at or below \subsubsection
\tableofcontents

\input{intro}

\input{3sf}

% Milestone 1 - Spec 1
\input{spec1}

% Milestone 2 - Spec 2
\input{spec2}

% Milestone 4 - Specs 3-4

\input{spec3}
\input{spec3-smt}
\input{spec3-alloy}
\input{spec4}
\input{spec4b}

\section{Summary}\label{sec:summary}
We have presented a series of specifications modeling the 3SF protocol from
various perspectives. Initially, we developed a direct translation of the
protocol's Python specification into \tlap{}, but this approach proved
unsatisfactory due to the reliance on recursion. To address this, we modified
the specification to use folds in place of recursion, theoretically enabling
model-checking with Apalache. However, this approach also proved impractical
due to the high computational complexity involved. Subsequently, we applied a
series of abstractions to improve the model-checking efficiency.

In addition to the \tlap{} specifications, we also introduced an SMT encoding
and an Alloy specification. The SMT encoding proved to be fairly performant,
while the Alloy specification demonstrated exceptional performance in
combination with the Kissat SAT solver.

To revisit the key outcomes of the project, see Section~\ref{sec:outcomes}.

\bibliographystyle{plain}
\bibliography{ref}

% appendix
\pagebreak

\appendix

The work done in this section is the main contribution of Milestone~3. Since
this section is quite long and technical, we have decided to add it in the
appendix.

% Milestone 3 - Translation rules
\input{rules}

\input{recursion-proofs}

\end{document}
